Concurrent Programming with CXL

### 2020-06-17 book/paper.tex
Started working on adding exercises to chapters

  description={is an atomic operation on the CXL machine, executed by a process}

@@

  description={is a variable that is stored in the memory of the CXL machine and

@@

  description={an assignment of values to variables.  In a CXL machine, this includes

@@

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapters~\ref{ch:cxlvalues} and~\ref{ch:cxlmachine} present some more
details of the CXL language and machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section, and Chapter~\ref{ch:naive} presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant implementation of a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\begin{quote}
Experiment!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares or factorials.
\end{itemize}
\end{quote}

@@

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will always end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

\begin{quote}
Can you think of a fix to the program?  You can try it out and see
if it works or not.
\end{quote}

\chapter{CXL Values}
\label{ch:cxlvalues}
\index{CXL Value}

CXL programs manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

\chapter{The CXL Machine}
\label{ch:cxlmachine}
\index{CXL machine}

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
    0 Jump 39
    1 Frame incrementer ['self'] 12
Up.cxl:2     count = count + 1;
    2 Push 1
    3 PushAddress count
    4 Load 1
    5 2-ary +
    6 PushAddress count
    7 Store 1
Up.cxl:3     done[self] = True;
    8 Push True
    9 PushVar self
   10 PushAddress done
   11 Store 2
   12 Return
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

@@

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the CXL machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}
a program counter,
\index{program counter}
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}
that contains a CXL value.
\end{itemize}

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
\glsadd{context}
it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
The CXL machine is predominantly a \emph{stack machine}.
\index{stack machine}
\glsadd{stack machine}
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process (with name tag \texttt{\_\_init\_\_/()}
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Once initialization completes, any processes that were spawned can run.
You can think of CXL as trying every possible interleaving of processes executing
instructions.
Figure~\ref{fig:incoutput} shows the output produced by running CXL on the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 9 in column 5 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
Next it reports an execution that failed this assertion.  The output has
four columns:
\begin{enumerate}
\item The name tag of the process;
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
\glsadd{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
\glsadd{state}
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
\glsadd{step}
Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
\glsadd{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and \texttt{x} is an arbitrary CXL value, then
\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
the same expression in CXL.

@@

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

@@

Chapter~\ref{ch:cxlvalues} provides an introduction to CXL values.
Below is a complete list of CXL value types with examples:

@@

\chapter{The CXL Machine}

The CXL machine has the following state:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of CXL machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionnary mapping atoms to values \\
\hline
ctxbag & a bag of non-stopped contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
failure & Boolean indicating some context has failed \\
\hline
initializing & Boolean indicating that the state is not fully initialized. \\
\hline
\end{tabular}
\vspace{1em}

@@

A \emph{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\index{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CXL machine
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.

>>>>

  description={is an atomic operation on the CXL virtual machine, executed by a process}

@@

  description={is a variable that is stored in the memory of the CXL virtual machine and

@@

  description={an assignment of values to variables.  In a CXL virtual machine, this includes

@@

\begin{itemize}
\item In Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} introduces the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter;
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a critical
section, and Chapter~\ref{ch:naive} presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant implementation of a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} introduces the notion of \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the bounded buffer
problem and various solutions.
\item Chapter~\ref{ch:starvation} talks about the problem that in some
synchronization solutions, processes may not be able to get access to a
resource they need, and a solution applied to the reader/writer problem.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present monitors and
condition variables, another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the actor model as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} shows solutions to process synchronization
that prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} introduces barrier synchronization, useful in
high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\section*{Exercises}
\begin{enumerate}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.)
\item Write a program that computes factorials.
\end{enumerate}

@@

\begin{quote}
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will always end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

\begin{quote}
\begin{itemize}
\item Can you think of a fix to the program?  You can try it out and see
if it works or not.
\end{itemize}
\end{quote}

\section*{Exercises}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/Up.py}
\end{code}
\caption{[Up.py] Python implementation of Figure~\ref{fig:inc}.}
\label{fig:incpy}
\end{figure}

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/UpMany.py}
\end{code}
\caption{[UpMany.py] Incrementing \texttt{N} times.}
\label{fig:incmany}
\end{figure}

\begin{enumerate}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:inc} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \texttt{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \texttt{count}
was for each of the failed runs.
\end{enumerate}

\chapter{The CXL Virtual Machine}
\label{ch:cxlmachine}
\index{CXL virtual machine}

CXL programs are compiled to CXL bytecode
\index{bytecode}
(machine instructions for a virtual machine),
which in turn is executed by the CXL virtual machine (CVM).
\index{virtual machine}
\index{CXL Virtual Machine}
\index{CVM}
To understand the problem of concurrent computing, it
is important to have a basic understanding of the CVM.

\section*{CXL Values}

CXL programs, and indeed the CVM,  manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

\section*{CXL Bytecode}

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
    0 Jump 39
    1 Frame incrementer ['self'] 12
Up.cxl:2     count = count + 1;
    2 Push 1
    3 PushAddress count
    4 Load 1
    5 2-ary +
    6 PushAddress count
    7 Store 1
Up.cxl:3     done[self] = True;
    8 Push True
    9 PushVar self
   10 PushAddress done
   11 Store 2
   12 Return
\end{verbatim}
\end{code}
\caption{The first part of the CVM bytecode corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

@@

A CXL program is translated
into \emph{bytecode}, which is list of machine instructions that the
CXL virtual machine (CVM) executes.
The CVM is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a CVM manipulates CXL values.
A CVM has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CVM instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CVM has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}
a program counter,
\index{program counter}
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}
that contains a CXL value.
\end{itemize}

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
\glsadd{context}
it contains the value of
its name tag, program counter, stack, and register.
The state of a CVM
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

@@

Compiling the code in Figure~\ref{fig:inc} results in the CVM bytecode
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
Each process in the CVM is predominantly a \emph{stack machine},
\index{stack machine}
\glsadd{stack machine}
but it also has a register.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process (with name tag \texttt{\_\_init\_\_/()}
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

Once initialization completes, any processes that were spawned can run.
You can think of CXL as trying every possible interleaving of processes executing
instructions.
Figure~\ref{fig:incoutput} shows the output produced by running CXL on the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 9 in column 5 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 9, 5) 1}.
Next it reports an execution that failed this assertion.  The output has
four columns:
\begin{enumerate}
\item The name tag of the process;
\item The sequence of program counters of the CVM instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

@@

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CVM bytecode.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CVM bytecode.  The first thing we need to
ensure is that there can never be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
\glsadd{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
\glsadd{state}
at any point of time the CVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
\glsadd{step}
Steps in CXL are atomic.

The CVM starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
\glsadd{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

@@

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlmachine}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and \texttt{x} is an arbitrary CXL value, then
\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
the same expression in CXL.

@@

While the CVM does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language in two ways.
First, any CXL statement can be made atomic by placing a label in front
of it.  Second, a group of CXL statements can be made atomic
through its \texttt{atomic}
\index{atomic statement}
statement.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc &(count);
\end{Verbatim}
\end{code}

@@

Chapter~\ref{ch:cxlmachine} provides an introduction to CXL values.
Below is a complete list of CXL value types with examples:

@@

\chapter{The CXL Virtual Machine}

The CXL Virtual Machine (CVM) has the following state:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of CVM machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionnary mapping atoms to values \\
\hline
ctxbag & a bag of non-stopped contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
failure & Boolean indicating some context has failed \\
\hline
initializing & Boolean indicating that the state is not fully initialized. \\
\hline
\end{tabular}
\vspace{1em}

@@

A \emph{micro step}
is the execution of a single CVM machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CVM can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

A \emph{macro step}
\index{macro step}
can involve multiple micro steps.  The following
instructions start a new macro step: \texttt{Load}, \texttt{Store},
\texttt{AtomicInc}, and \texttt{Continue}.  The CVM
interleaves macro steps, not micro steps.  Like micro steps, each
macro step involves a single context.  Unlike a micro step, a macro
step can leave the state unchanged.
